---
title: "Practicals: Analysis of Count Data"
author: |
  | Kamarul Imran Musa
  | Assoc Prof (Epidemiology and Statistics
date: "`r Sys.Date()`"
output:
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: 3
  html_document:
    highlight: tango
    number_sections: yes
    theme: united
    toc: yes
    toc_depth: '3'
    toc_float: yes
---

\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Prepare environment

Load libraries

```{r}
library(tidyverse)
library(haven)
library(broom)
library(janitor)
```


# Data

## Read data

Two datasets for practice

```{r}
publish <- read_stata('couart2.dta')
visit <- read_csv('NMES1988.csv') %>% clean_names()
```

## Transform data

Take a peek of data

```{r}
glimpse(publish)
```


For variables:

- fem
- mar

```{r}
publish <- publish %>% mutate_if(is.labelled, ~as_factor(.))
```


# Describe data

## Summarize data

Let us see summary of data

```{r}
summary(publish)
```

The outcome variable is **art** : that is the number of publications produced by PhD biochemists. 

## Plots outcome variable

```{r}
ggplot(publish, aes(x = art)) + geom_histogram() 
```


# Estimation

Let us assume the outcome follows the Poisson regression

- random observations
- independent observations 
- lineariry in parameters
- mean = variance 

## Constant only model

Let us estimate a constant model only. No covariates.

```{r}
mod_0 <- glm(art ~ 1, family = poisson(link = 'log'), 
             publish)
summary(mod_0)
```

The estimated parameters; $\beta_0$ = `r coef(mod_0)`, hence $\mu$ = `r exp(coef(mod_0))` (The estimated mean of `art`). This equals the mean of `art` that is `r mean(publish$art)`.

## Multivariable model

```{r}
mod_1 <- glm(art ~ fem + mar + kid5 + phd + ment, family = poisson(link = 'log'), 
             data = publish)
summary(mod_1)
```

Or you can use `tidy()` for nicer output

```{r}
tidy(mod_1, conf.int = TRUE)
```

## Interpretation

The expected log of count

- the number of articles for women was on average -0.225 the log of article number compared to men
- the number of articles for married scholar was on average 0.155 the log of article number compared to single scholars

The expected rate:

If we want to interpret using the rate, $\mu$, then for a unit change in $x_k$, the expected count changes by a factor of $exp^{\beta_k}$, holding other other variables constant.  

```{r}
tidy(mod_1, exponentiate = TRUE, conf.int = TRUE)
```

So, being a female scientist ($x_{fem}$) decreases the expected number of articles (**art**) by a factor of `r exp(coef(mod_1)[2])`, holding other other variables constant.  

# Prediction

## Predicted log of count

The predicted log of articles for **art**

```{r}
augment(mod_1)
```

## Predicted count

The predicted count of **art** for the given covariates for each subject in the data 

```{r}
exp_count <- predict(mod_1, type = 'response')
head(exp_count)
```

Manually, for the 1st subject, we calculate

```{r}
# The predicted log of count
0.30461683 - 0.22459423*0 +  0.15524338*1 - 0.18488270*0 + 0.01282258*2.52 + 0.02554275*7
# The expected (average) count of art 
exp(0.6709724)
```

# References

1.  Long J S, Freese J. Regression Models for Categorical Dependent Variables Using Stata
2. https://cran.r-project.org/web/packages/AER/AER.pdf 

